---
title: "Who is Fueling the 2025 New Jersey Governor Race?"
output: html_document
date: "2025-06-02"
---

Levi Jiang
lfjiang@uw.edu

<b>What is the story: </b>

The 2025 election for New Jersey’s Governor is getting intense. The New York Times has described it as “The race features the largest field in decades. The candidates have never been more accomplished or better funded. The results could be a bellwether for midterm congressional elections.”  

New Jersey is one of only two states (another one is Virginia) that holds a governor election in the year following a presidential election. For Democrats, it could reveal if they are winning back New Jersey voters who shifted right in the 2024 election. This makes the election an indicator of the popularity of the current Trump administration and the midterm elections in 2026. 

As the race heats up, my story will examine the source of each candidate's election funds, specifically, their donations. This article will disclose the identities of donors, and explore whether there are any patterns worth noting.


<b>Why NOW: </b>

On June 10th, Six Democrats and five Republicans will compete in the primary on June 10th for their party’s nomination, and the winners will advance to the final election in November. The primaries are bound to attract public attention, and this is the right time to inform readers about the candidates' campaign finances.


<b>Target audience:</b>

1. New Jersey residents who want to know more about their governor candidates
2. Election stakeholders (e.g., campaign staff, donors, potential donors)
3. Politicians, political analysts, commentators, enthusiasts (especially national politicians to see if political trends change following the Democrats' wipeout in the 2024 elections)


<b>Potential impact:</b>
We want to help the readers deepen their understanding of this election and each candidate, motivate more people to vote, and facilitate their decision-making.


<b>Questions I want to answer:</b>

1. Who are the biggest donors to each candidate? How much did they donate? Where are they from? What industry do they work in? If we can find some connections between candidates and executives from health care corporations, union members etc, we can run deeper investigations on that.
2. Do certain candidates receive more donations from specific industries because of their special political stances? (This can also be used to explore geographic disparities: are there candidates who see donation advantages in certain neighborhoods?)
3. Are there out-state contributors making significant impacts on the election? (And also for in-state donors, we can use their zip to find some patterns.)


<b>Data Visualization (according to the questions I pitched):</b>

1. A chart to show everyone’s top 10 donors and their info
2. A pie chart or bar chart to show which industries the donations come from.
3. A US map to show which states the donations are from. For in-state donations, we can use a NJ State map and show the areas that donations are from based on zip codes.


<b>Data Sources:</b>

1. New Jersey ELEC Reports and Data Search System - candidate’s donation data (https://www.njelecefilesearch.com/searchcandidatereports)
2. Ballot Pedia - Info of each candidate (https://ballotpedia.org/New_Jersey_gubernatorial_and_lieutenant_gubernatorial_election,_2025)
3. New Jersey Globe - the latest headlines and results
(https://newjerseyglobe.com/)


<b>Previous Coverage:</b>

There’s not so much data analysis about this election. The New York Times has an article about who the candidates are and their political stances: 
(https://www.nytimes.com/2025/06/02/nyregion/new-jersey-governors-race-what-to-know.html)
(https://www.nytimes.com/interactive/2025/03/21/nyregion/new-jersey-governor-candidates-issues.html)

Politico also analyzed everyone’s advantages and disadvantages:
(https://www.politico.com/interactives/2025/nj-governor-candidates-comparison-analysis/)

Ballotpedia made a map of the results of New Jersey county Democratic Party conventions: 
(https://datawrapper.dwcdn.net/Ho3TT/3/)


<b>Other information:</b>

At this point we know that there will be 11 candidates running for the Governor. Thoroughly analyzing all of them in a week will be difficult. Rob and I shared a great idea: analyze only the top five candidates in the polls. The idea of analyzing the rest of the candidates is the same as them, as is the visualization process. 

According to the newest Emerson College Polling/PIX11/The Hill survey, the top 5 candidates are: Mikie Sherrill (D), Jack Ciattarelli (R), Bill Spadea (R), Steven Fulop (D), and Ras Baraka (D).



# Load software libraries
```{r}

#install.packages("tidyverse")
#install.packages("janitor")
#install.packages("readxl")
#install.packages("rvest")
#install.packages("dplyr")
library(tidyverse)
library(janitor)
library(readxl)
library(rvest)
library(dplyr)

```


# Load data
```{r}

sherrill <- read.csv("../data/sherrill_mikie_Cont_638821188859939969.csv") |> 
    clean_names() 
sweeney <- read.csv("../data/sweeney_stephen_Cont_638821191659704440.csv") |> 
    clean_names() 


```

Data details:

# Explore the data types
```{r}

combo <- rbind(sherrill, sweeney)
glimpse(sherrill)

```
# Fix dates
```{r}
sherrill <- sherrill |> 
  mutate(date = lubridate::mdy(contribution_date))
sweeney <- sweeney |> 
  mutate(date = lubridate::mdy(contribution_date))
```


# Data exploration: totals by candidate

```{r}
combo <- rbind(sherrill, sweeney)

combo |> 
  select(entity_name, contribution_amount) |> 
  group_by(entity_name) |> 
   summarize(total = sum(contribution_amount)) |> 
  arrange(desc(total))
```

So we can see 

# Questions --------------------------------------------------------------------

Write down questions you want answered:

First of all we need to clean the dataset. The original spreadsheet mix the individual donors and organizations, so we need to create an id for each donor. Also, some donors made more than one donation or changes the amount of money, then we need to figure out the exact numbers.

```{r}
# Data Clean Up

# Step 1: Create a donor ID (individual name or organization name)
sherrill_donations_clean <- sherrill %>%
  mutate(
    donor_id = case_when(
      !is.na(first_name) & first_name != "" &
      !is.na(last_name) & last_name != "" ~ str_trim(paste(first_name, last_name)),
      !is.na(non_ind_name) & non_ind_name != "" ~ str_trim(non_ind_name),
      TRUE ~ "Unknown"
    )
  )

# Step 2: Group by donor_id and sum up all contribution_amount
sherrill_donor_totals <- sherrill_donations_clean %>%
  group_by(donor_id) %>%
  summarise(final_donation = sum(contribution_amount, na.rm = TRUE), .groups = "drop")

# Step 3: Merge the final total back to the original dataset
sherrill_donations_with_total <- sherrill_donations_clean %>%
  left_join(sherrill_donor_totals, by = "donor_id")

# Step 4: Select relevant columns from the original data
sherrill_donor_details <- sherrill_donations_with_total %>%
  select(
    donor_id, final_donation, is_individual, date,
    street, city, state, zip,
    emp_name, emp_street, emp_city, emp_state, emp_zip,
    occupation_name, contributor_type
  )

# Step 5: Keep only the newest row for each donor based on date
# Convert date to proper Date type if it's not already
sherrill_donor_details <- sherrill_donor_details %>%
  mutate(date = as.Date(date))

sherrill_latest_details <- sherrill_donor_details %>%
  arrange(donor_id, desc(date)) %>%
  group_by(donor_id) %>%
  slice(1) %>%  # Keep only the newest record
  ungroup()


# So as for other candidates

sweeney_donations_clean <- sweeney %>%
  mutate(
    donor_id = case_when(
      !is.na(first_name) & first_name != "" &
      !is.na(last_name) & last_name != "" ~ str_trim(paste(first_name, last_name)),
      !is.na(non_ind_name) & non_ind_name != "" ~ str_trim(non_ind_name),
      TRUE ~ "Unknown"
    )
  )

sweeney_donor_totals <- sweeney_donations_clean %>%
  group_by(donor_id) %>%
  summarise(final_donation = sum(contribution_amount, na.rm = TRUE), .groups = "drop")

sweeney_donations_with_total <- sweeney_donations_clean %>%
  left_join(sweeney_donor_totals, by = "donor_id")

sweeney_donor_details <- sweeney_donations_with_total %>%
  select(
    donor_id, final_donation, is_individual, date,
    street, city, state, zip,
    emp_name, emp_street, emp_city, emp_state, emp_zip,
    occupation_name, contributor_type
  )

sweeney_donor_details <- sweeney_donor_details %>%
  mutate(date = as.Date(date))

sweeney_latest_details <- sweeney_donor_details %>%
  arrange(donor_id, desc(date)) %>%
  group_by(donor_id) %>%
  slice(1) %>%  # Keep only the newest record
  ungroup()


```

<b>Q1: Who are the biggest donors to each candidate? How much did they donate? Where are they from? What industry do they work in? (If we can find some connections between candidates and executives from health care corporations, union members etc, we can run deeper investigations on that.)</b>

So for candidate Mikie Sherrill:

```{r}

sherrill_top_donations <- sherrill_latest_details %>%
  arrange(desc(final_donation)) %>%
  slice_head(n = 200)
print(sherrill_top_donations)

```
 
 As we can see, the money they donated are all the same, so it's hard to tell who's the "biggest" boss. We can partly sort them out by whether they are individual or not. 
 (What does a journalist usually do when this happens?)
 
```{r}

sweeney_top_donations <- sweeney_latest_details %>%
  arrange(desc(final_donation)) %>%
  slice_head(n = 20)
print(sherrill_top_donations)

```

```{r}

sweeney_top_donations <- sweeney_latest_details %>%
  arrange(desc(final_donation)) %>%
  slice_head(n = 200)
print(sherrill_top_donations)

```

donations_money_by_state_sherrill <- sherrill %>%
  group_by(state) %>%
  summarise(Total = sum(contribution_amount), .groups = "drop") %>%
  arrange(desc(Total))

donations_amount_by_state_sherrill <- sherrill %>%
  group_by(state) %>%
  summarise(Count = n(), .groups = "drop") %>%
  arrange(desc(Count))






donations_money_by_state_sherrill <- sherrill %>%
  group_by(state) %>%
  summarise(Total = sum(contribution_amount), .groups = "drop") %>%
  arrange(desc(Total))

donations_amount_by_state_sherrill <- sherrill %>%
  group_by(state) %>%
  summarise(Count = n(), .groups = "drop") %>%
  arrange(desc(Count))



